Internal models of type theory

### 2013-11-01 article-models-main.tex
moved content to the new file
funcy identity types
compatibility rules for weakening
developing category theory inside type theory; other constructors

@@

>>>>

\newcommand{\eqtype}[1]{\typefont{Eq}_{#1}}
\newcommand{\eqtoid}[1]{\typefont{eqtoid}(#1)}
\newcommand{\greek}{\mathrm}
\newcommand{\product}[2]{{#1}\times{#2}}
\newcommand{\pairp}[1]{(#1)}
\newcommand{\ctxwkop}[2]{\ctxwk{#2}{#1}}
\makeatletter
\newcommand{\jhom}[3]{\jterm{\jhom@dom@ctx #1}{\jhom@dom@wk{#1}{#2}}{#3}}
\newcommand{\jhom@dom@ctx}{\@ifnextchar\bgroup{\@jhom@dom@ctx}{}}
\newcommand{\@jhom@dom@ctx}[1]{\@ifnextchar\bgroup{\@@jhom@dom@ctx{#1}}{#1}}
\newcommand{\@@jhom@dom@ctx}[2]{\ctxext{\jhom@dom@ctx #1}{\jhom@dom@wk{#1}{#2}}}
\newcommand{\jhom@dom@wk}[2]{{\@ifnextchar\bgroup{\@jhom@dom@wk{#2}}{\ctxwkop{#2}\bgroup\aftergroup{\aftergroup{\egroup}}}#1}}
\newcommand{\@jhom@dom@wk}[2]{\@ifnextchar\bgroup{\@@jhom@dom@wk{#2}{#1}}{\ctxwk{#1}{#2}}}
\newcommand{\@@jhom@dom@wk}[3]{\jhom@dom@wk{#2}{\ctxwk{\jhom@dom@ctx #3}{#1}}}
\makeatother
\newcommand{\jhomeq}[4]{\jtermeq{#1}{\ctxwk{#1}{#2}}{#3}{#4}}
\newcommand{\jcomp}[3]{\subst{#2}{\ctxwk{#1}{#3}}}
\newcommand{\jequalizer}[3]{\{#1|#2\jdeq #3\}}
\newcommand{\jequalizerin}[2]{\iota_{#1,#2}}
\newcommand{\tounit}[1]{{!_{#1}}}
\newcommand{\trwk}{\typefont{trwk}}
\newcommand{\trext}{\typefont{trext}}

@@

\input{article-models-tt}

### 2013-11-01 article-models-tt.tex
developing category theory inside type theory; other constructors

\section{A variable-free presentation of type theory}
We have seen in the example of $\UU$ that we will not need to refer to variables
when we are manipulating contexts. In fact, it is more natural not to.
This suggests that there is a presentation
of type theory where contexts do not mention variables. The first small gain is that such a presentation
would not be burdened with comments about variables being bounded or not, or fresh or
not occuring at all.

@@

@@

@@

\subsection{Other type constructors}
We will now describe the rules for the type constructors that we don't assume
in our type theory by default. The first of these are the dependent function
types. We also discuss $\tfW$-types, suspensions, truncations. More general higher inductive
types will have to wait until we have introduced models, for the models of the
basic type theory because we will use them as index categories of the diagrams.

\subsubsection{Dependent function types}
\begin{infarray}{cc}
\inference{\jtype{\ctxext{\Gamma}{A}}{P}}{\jtype{\Gamma}{\mprd{A}{P}}}
& \inference{\jtypeeq{\ctxext{\Gamma}{A}}{P}{P'}}{\jtypeeq{\Gamma}{\mprd{A}{P}}{\mprd{A}{P'}}}\\
\inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jterm{\Gamma}{\mprd{A}{P}}{\lambda(f)}}
& \inference{\jtermeq{\ctxext{\Gamma}{A}}{P}{f}{f'}}{\jtermeq{\Gamma}{\mprd{A}{P}}{\lambda(f)}{\lambda(f')}}\\
\inference{\jterm{\Gamma}{\mprd{A}{P}}{g}}{\jterm{\ctxext{\Gamma}{A}}{P}{\tfev(g)}}
& \inference{\jtermeq{\Gamma}{\mprd{A}{P}}{g}{g'}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\tfev(g)}{\tfev(g')}}\\
\inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\Gamma}{A}{\tfev(\lambda(f))}{f}}
\end{infarray}

@@

>>>>

\section{Variable-free type theory -- the basic theory}
We have seen in the example of $\UU$ that we will not need to refer to variables
when we are manipulating contexts. In fact, it is more natural not to.
This suggests that there is a presentation
of type theory where contexts do not mention variables. The first small gain is that such a presentation
would not be burdened with comments about variables being bounded or not, or fresh or
not occuring at all.

@@

\subsubsection{A game of trees}
We prove a little meta theorem about the type system we have so far. It is not
really of importance to the development of the theory.

Let's say
that a binary tree of contexts is a pair $\pair{T,f}$ consisting of a binary 
tree $T$ together with a function $f$ assigning to each leaf a context. The set
of all binary trees of contexts is denoted by $B$. Such
binary trees may be defined inductively: $(\unit,f):B$ for any $f:\unit\to ctx$
and given any $(T_1,f_1)$ and $(T_2,f_2)$ in $B$ we have $(T_1,f_1)*(T_2,f_2)$
in $B$.

We simultaneously define
the following two functions:
\begin{align*}
\trext & : B\to ctx\\
\trwk_0 & : \prd{X:B} ctx\to typ(\trext(X))\\
\trwk_1 & : \prd{X:B}{\Gamma~ctx} typ(\Gamma)\to typ(\ctxext{\trext(X)}{\trwk_0(X,\Gamma)})
\end{align*}
Both these functions are defined by induction on binary trees. We set
\begin{align*}
\trext((\unit,f)) & \defeq f(\unit)\\
\trext((T_1,f_1)*(T_2,f_2)) & \defeq \ctxext{\trext((T_1,f_1))}
{\trwk_0((T_1,f_1),\trext((T_2,f_2)))}
\end{align*}
and for any context $\Gamma$
\begin{align*}
\trwk_0((\unit,f),\Gamma) & \defeq \ctxwk{f(\unit)}{\Gamma}\\
\trwk_0((T_1,f_1)*(T_2,f_2),\Gamma) 
  & \defeq \trwk_1((T_1,f_1),\trwk_0((T_2,f_2)),\Gamma))
\end{align*}
and for any type $A$ in context $\Gamma$
\begin{align*}
\trwk_1((\unit,f),A) & \defeq \ctxwk{f(\unit)}{A}\\
\trwk_1((T_1,f_1)*(T_2,f_2),A) & \defeq
  \trwk_1((T_1,f_1),\trwk_1((T_2,f_2)),A))
\end{align*}

\begin{lem}
$\trwk_0((T,f),\Gamma)$ is a type in context $\trext((T,f))$ and
$\trwk_1((T,f),A)$ is a type in context $\ctxext{\trext((T,f))}{\trwk_0((T,f),\Gamma)}$
for any type $A$ in context $\Gamma$.
\end{lem}

\begin{proof}
It is immediate that $\trwk_0((\unit,f),\Gamma)$ is a type in context $\trext((\unit,f))$
and that $\trwk_1((\unit,f),A)$ is a type in context $\ctxext{\trext((\unit,f))}
{\trwk_0((\unit,f),\Gamma)}$ for any type $A$ in context $\Gamma$.

Suppose that $(T_1,f_1)$ and $(T_2,f_2)$ are binary trees of contexts such that
$\trwk_0((T_i,f_i),\Gamma)$ is a type in context
$\trext((T_i,f_i))$ for $i\jdeq 1$ and $i\jdeq 2$, and such that
$\trwk_1((T_i,f_1),A)$ is a type in context $\ctxext{\trext((T_i,f_i))}{\trwk_0((T_i,f_i),\Gamma)}$
for $i\jdeq 1$ and $i\jdeq 2$. Then
\begin{equation*}
\trwk_0((T_1,f_1)*(T_2,f_2),\Gamma) 
\jdeq \trwk_1((T_1,f_1),\trwk_0((T_2,f_2),\Gamma))
\end{equation*}
is a type in context
\begin{equation*}
\ctxext{\trext((T_1,f_1))}{\trwk_0((T_1,f_1),\trext((T_2,f_2)))}
\jdeq \trext((T_1,f_1)*(T_2,f_2)).
\end{equation*}
Also,
\begin{equation*}
\trwk_1((T_1,f_1)*(T_2,f_2),A) \jdeq \trwk_1((T_1,f_1),\trwk_1((T_2,f_2),A))
\end{equation*}
is a type in context
\begin{equation*}
\ctxext{\trext((T_1,f_1))}{\trwk_0((T_1,f_1),\ctxext{\trext((T_2,f_2))}{\trwk_0((T_2,f_2),\Gamma))}}
\end{equation*}
\end{proof}

@@

Using the rules of the compatibility of substitution with weakening and of the
compatibility of weakening with itself, we see that we can show
\begin{equation*}
\jtypeeq{\Gamma}{\subst{f}{\ctxwk{\Gamma}{{\Delta}{\greek{E}}}}}{\ctxwk{\Gamma}{E}}
\end{equation*}
for any three contexts $\Gamma$, $\Delta$ and $\greek{E}$ and any $\jhom{\Gamma}{\Delta}{f}$.
It follows that for $\jhom{\Delta}{\greek{E}}{g}$ we have
\begin{equation*}
\jhom{\Gamma}{\greek{E}}{\subst{f}{\ctxwk{\Gamma}{g}}}
\end{equation*}
The term $\jcomp{\Gamma}{f}{g}$ is the composition of $g$ with $f$.

@@

\section{Variable-free type theory -- other type constructors}
We will now describe the rules for the type constructors that we don't assume
in our type theory by default. The first of these are the dependent function
types. We also discuss $\tfW$-types, suspensions, truncations. More general higher inductive
types will have to wait until we have introduced models, for the models of the
basic type theory because we will use them as index categories of the diagrams.

\subsection{The unit type}
Since we don't have a notion of terms of a context, we just say that the context
$\unit$ is the terminal context.

\begin{infarray}{c}
\inference{}{\jctx{\unit}}\\
\inference{\jctx{\Gamma}}{\jhom{\Gamma}{\unit}{\tounit{\Gamma}}}\\
\inference{\jctx{\Gamma}\qquad\jhom{\Gamma}{\unit}{f}}{\jhomeq{\Gamma}{\unit}{f}{\tounit{\Gamma}}}
\end{infarray}

Note that we don't have to require that $\jhomeq{\Gamma}{\unit}{\tounit{\Gamma}}{\tounit{\Delta}}$
whenever we have a judgmental equality $\jctxeq{\Gamma}{\Delta}$, since this already follows from the third rule.

When the context $\unit$ is present, we may use the expression $\jtype{}{\Gamma}$
as a shorthand for the judgment $\jtype{\unit}{\ctxwk{\unit}{\Gamma}}$. Likewise,
we may use the expression $\jtermc{\Gamma}{i}$ as a shorthand
for the judgment $\jhom{\unit}{\Gamma}{i}$. If we have a type $A$ in context
$\Gamma$, we may use the expression $\jtype{}{\subst{i}{A}}$ to stand for
the judgment $\jtype{\unit}{\subst{i}{\ctxwk{\unit}{A}}}$. We see that in every
respect, contexts are types in the empty context.

We have created a strictly terminal object $\unit$. This is not necessary when
we're working in a context. We introduce the unit type $\unit_\Gamma$ in context
$\Gamma$ in the familiar type theoretical way.

\begin{infarray}{c}
\inference{\jctx{\Gamma}}{\jtype{\Gamma}{\unit_\Gamma}}\\
\inference{\jctx{\Gamma}}{\jterm{\Gamma}{\unit_\Gamma}{\ttt_\Gamma}}\\
\inference{\jtype{\ctxext{\Gamma}{\unit_\Gamma}}{P}\qquad\jterm{\Gamma}{\subst{\ttt_\Gamma}{P}}{u}}
          {\jterm{\ctxext{\Gamma}{\unit_\Gamma}}{P}{\ind{\unit_\Gamma}(u)}}\\
\inference{\jtype{\ctxext{\Gamma}{\unit_\Gamma}}{P}\qquad\jterm{\Gamma}{\subst{\ttt_\Gamma}{P}}{u}}
          {\jtermeq{\Gamma}{\subst{\ttt_\Gamma}{P}}{\subst{\ttt_\Gamma}{\ind{\unit_\Gamma}(u)}}{u}}
\end{infarray}

\subsection{Product types}

\subsubsection{Products}
\begin{infarray}{c}
\inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jctx{\product{\Gamma}{\Delta}}}\\
\inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jhom{{\Gamma}{\Delta}}{\product{\Gamma}{\Delta}}{\pair}}
\end{infarray}

\subsubsection{Strict products}
As with the unit type, we may use the categorical description of the product
for our type theoretical definition. If we do that, we get strict products.

\begin{infarray}{cc}
\inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jctx{\product{\Gamma}{\Delta}}}
 & \inference{\jhom{\greek{E}}{\Gamma}{f}\qquad\jhom{\greek{E}}{\Delta}{g}}
          {\jhom{\greek{E}}{\product{\Gamma}{\Delta}}{\pairp{f,g}}}\\
\inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jhom{\product{\Gamma}{\Delta}}{\Gamma}{\proj1}}
 & \inference{\jhom{\greek{E}}{\Gamma}{f}\qquad\jhom{\greek{E}}{\Delta}{g}}
          {\jhomeq{\greek{E}}{\Gamma}{\jcomp{\greek{E}}{\pairp{f,g}}{\proj1}}{f}}\\
\inference{\jctx{\Gamma}\qquad\jctx{\Delta}}{\jhom{\product{\Gamma}{\Delta}}{\Delta}{\proj2}} 
 & \inference{\jhom{\greek{E}}{\Gamma}{f}\qquad\jhom{\greek{E}}{\Delta}{g}}
          {\jhomeq{\greek{E}}{\Delta}{\jcomp{\greek{E}}{\pairp{f,g}}{\proj2}}{g}}
\end{infarray}
\begin{equation*}
\inference{\jhom{\greek{E}}{\product{\Gamma}{\Delta}}{h}
           \qquad
           \jhomeq{\greek{E}}{\Gamma}{\jcomp{E}{h}{\proj1}}{f}
           \qquad
           \jhomeq{\greek{E}}{\Delta}{\jcomp{E}{h}{\proj2}}{g}}
          {\jhomeq{\greek{E}}{\product{\Gamma}{\Delta}}{h}{\pairp{f,g}}}
\end{equation*}

\subsection{Equalizer types}
Now that we have introduced a terminal object and products the categorical way,
we may just continue and present (strict) equalizers and pullbacks too, just to
see where we get. These notions are probably just not very useful in a univalent
setting until we got a good computational interpretation.

\begin{infarray}{c}
\inference{\jhom{\Gamma}{\Delta}{f}\qquad\jhom{\Gamma}{\Delta}{g}}{\jctx{\jequalizer{\Gamma}{f}{g}}}\\
\inference{\jhom{\Gamma}{\Delta}{f}\qquad\jhom{\Gamma}{\Delta}{g}}{\jhom{\jequalizer{\Gamma}{f}{g}}{\Gamma}{\jequalizerin{f}{g}}}\\
\inference{\jhom{\greek{E}}{\Gamma}{h}\qquad\jhomeq{\greek{E}}{\Delta}{\jcomp{\greek{E}}{h}{f}}{\jcomp{\greek{E}}{h}{g}}}{\jhom{\greek{E}}{\jequalizer{\Gamma}{f}{g}}{\jequalizer{h}{f}{g}}}\\
\inference{
  {\begin{array}{l}
    \jhom{\greek{E}}{\Gamma}{h}\\
    \jhomeq{\greek{E}}{\Delta}{\jcomp{\greek{E}}{h}{f}}{\jcomp{\greek{E}}{h}{g}}
  \end{array}}
  \qquad
  {\begin{array}{l}
    \jhom{\greek{E}}{\jequalizer{\Gamma}{f}{g}}{k}\\
    \jhomeq{\greek{E}}{\Gamma}{\jcomp{\greek{E}}{k}{\jequalizerin{f}{g}}}{h}
  \end{array}}
}
  {\jhom{\greek{E}}{\jequalizer{\Gamma}{f}{g}}{\jequalizer{h}{f}{g}}}
\end{infarray}

\subsection{Pullback types}

\subsection{Dependent function types}
\begin{infarray}{cc}
\inference{\jtype{\ctxext{\Gamma}{A}}{P}}{\jtype{\Gamma}{\mprd{A}{P}}}
& \inference{\jtypeeq{\ctxext{\Gamma}{A}}{P}{P'}}{\jtypeeq{\Gamma}{\mprd{A}{P}}{\mprd{A}{P'}}}\\
\inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jterm{\Gamma}{\mprd{A}{P}}{\lambda(f)}}
& \inference{\jtermeq{\ctxext{\Gamma}{A}}{P}{f}{f'}}{\jtermeq{\Gamma}{\mprd{A}{P}}{\lambda(f)}{\lambda(f')}}\\
\inference{\jterm{\Gamma}{\mprd{A}{P}}{g}}{\jterm{\ctxext{\Gamma}{A}}{P}{\tfev(g)}}
& \inference{\jtermeq{\Gamma}{\mprd{A}{P}}{g}{g'}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\tfev(g)}{\tfev(g')}}\\
\inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\Gamma}{A}{\tfev(\lambda(f))}{f}}
\end{infarray}

@@

\subsection{The empty type}

\subsection{Coproduct types}

\subsection{The natural numbers}
